An Algebraic Representation of Wireless Workload

### 2008-02-08 paper.tex
Finalized through the common representations section.

@@

The simplifying assumptions made by most traffic models can be classified into three categories. We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions. They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows.

@@

Since uniform behaviors entail homogeneity of properties with respect to each single flow individually, a trace matrix can be transformed to the closest matrix satisfying the model's assumptions via right matrix multiplication. Right multiplication mixes the elements within each row, and if the mixing weights are all equal, the results is homogenized behavior. For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
\begin{align*}
S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
\end{align*}

@@

It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees. This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives. Here we examine how various notions of uniformity are expressed in the matrix model. In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows. The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies. The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}. The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
\begin{align}
\Gamma=\begin{pmatrix}
\ones{} & \vec{\sigma} & \zeros{} \\
\zeros{} & \zeros{} & \vec{\tau}
\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
\end{align}

While constant bit-rate in the strict sense implies only that the packet behavior in each flow is uniform, it is very rare for it to be applied in this strict sense only: it is typical that ``\caps{CBR}'' models make assumptions about uniformity across flows, and not just within each flow. The most drastic simplification is to treat all flows in a synthetic workload identically. This is effected by left multiplication: $F'=(\ones{f \times f}/f)F$. To make flow behavior uniform on a per-source-node basis, we left-multiply the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$. Per-destination-node uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.

Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified. The confusion arises due to the lack of a unique way of associating flows with nodes. Each node can equally viably be associated with its source or its destination. Which association is more valid? Consider a web request: should the behavior of the request be determined by what the client typically does? Or should the requests that the server typically receives determine behavior? There is no armchair answer: only analysis of real data can give us the answer. The matrix model provides an analytical framework for doing so.

The matrix model has the benefit of making the uniformization of behaviors completely explicit. It also allows us to easily define intermediate choices. If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
\begin{align}
F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
\end{align}
The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors. Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.

@@

Marginality is an orthogonal concept to uniformity: it implies that the distribution of some property is identical \textit{across} all flows. For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size. The traffic models proposed by {\FHC}~\cite{Hernandez06:wlan-traffic} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node. These models do not account for non-independence between these and other flow properties. We will discuss this limitation and its implications when we analyze our experimental results.

\subsubsection{Regularized Behavior}

The final form of sameness we consider, is regularization. This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings. For example, we can provide a distribution of flow sizes on a per source-node basis; that is, each source node can have it's \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.

@@

\subsection{Uniformity}

\begin{figure*}[tb]
\begin{center}
\subfloat[trace node behavior]
{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[uniform node behavior]
{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[trace flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[uniform flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[packet behavior]
{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels. Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway). Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Figure \ref{fig:packet-behavior} compares uniform (i.e. {\footnotesize{CBR}}) packet behavior with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
\label{fig:trace-vs-uniform}
\end{center}
\vspace{-2em}
\end{figure*}

% TODO: differentiate between restrictive versions of models and transformative versions.
% TODO: distinguish uniform, marginal and regularized models.

\subsection{Marginality}

Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior. For example, {\FHC}~\cite{Hernandez06:wlan-traffic} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution. This is certainly much greater detail than simply assuming that all nodes and flows are identical.

@@

\section{Experimental Methodology}\label{sec:methodology}

@@

\begin{table*}
\begin{center}
\small
\begin{tabular}{|c|c|l|}
\multicolumn{1}{c}{\textbf{model}} &
\multicolumn{1}{c}{\textbf{transformation}} &
\multicolumn{1}{c}{\textbf{description}} \\
\hline
{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline

time uniform & $A'=A(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
node uniform & $S'=S(\ones{n \times n}/n)$~~~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline
\hline
time marginal & $A'=(\ones{f \times f}/f)A$ & start time independently chosen from overall distribution of start times \\\hline
size marginal & $B'=(\ones{f \times f}/f)B$ & flow size (bytes) is chosen independently of other properties \\\hline
node marginal & $S'=(\ones{f \times f}/f)S$~~~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
packet marginal & $Z'=(\ones{f \times f}/f)Z$~~~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline
\hline
source regular & $F'=(SS^t/f)F$ & flow behaviors are aggregated on a per source-node basis \\\hline
dest regular & $F'=(DD^t/f)F$ & flow behaviors are aggregated on a per destination-node basis \\\hline
mixed regular & $F'=\frac{1}{2f}(SS^t+DD^t)F$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline

\end{tabular}
\caption{Matrix-based traffic models evaluated by paired differential simulation.}
\label{tab:traffic-models}
\end{center}
\end{table*}

@@

\begin{figure}
\vspace{0.45em}
\begin{center}
\includegraphics[width=3.5in]{nodes-flows}%
\vspace{-0.75em}%
\caption{The number of active nodes and flows over time.} 
\label{fig:nodes-flows}
\end{center}
\vspace{-2em}
\end{figure}

>>>>

\begin{figure*}[tb]
\begin{center}
\subfloat[trace node behavior]
{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[uniform node behavior]
{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[trace flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[uniform flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[packet behavior]
{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
\caption{Examples illustrating trace (real) versus uniform (na\"ive synthetic) behaviors at the node, flow and packet levels. Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway). Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Figure \ref{fig:packet-behavior} compares uniform (i.e. {\footnotesize{CBR}}) packet behavior with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
\label{fig:trace-vs-uniform}
\end{center}
\vspace{-1.25em}
\end{figure*}

@@

The simplifying assumptions made by most traffic models can be classified into three categories. We use the terms \textit{uniformity}, \textit{marginality}, and \textit{regularity} do describe these classes of assumptions. They are all assumptions about the ``sameness'' of aspects of network behaviors; they differ, however, in whether the make stipulations about uniformity within each flow, across all flows, or within groups of flows. In the following sections, we explore how various modeling concepts can be expressed in the framework of the General Matrix Model. Models are expressed as transformations that will convert arbitrary traffic matrix instances into similar ones that satisfy the model's assumptions. These transformations will later allow us to evaluate the effect of each model's simplifications on the realism of generated workloads.

@@

Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication. Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior. For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
\begin{align*}
S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
\end{align*}
This makes all the values in each flow's row identically the average of the original $n$ values. As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies. The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}. The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
\begin{align}
\Gamma=\begin{pmatrix}
\ones{} & \vec{\sigma} & \zeros{} \\
\zeros{} & \zeros{} & \vec{\tau}
\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
\end{align}
\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to actual behaviors from traces.
%By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition

@@

%It is common in generating traffic behaviors for experiments to treat packets, nodes, and flows uniformly to varying degrees. This is largely because of the high complexity of doing otherwise and the lack of compelling alternatives. Here we examine how various notions of uniformity are expressed in the matrix model. In general, \textit{intra-flow uniformity} is effected by right multiplication, which mixes the elements in each row, whereas \textit{inter-flow uniformity} is effected by left multiplication, which mixes elements across rows. The constant bit-rate (\caps{CBR}) flow model, for example, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies. The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}. The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
%\begin{align}
%\Gamma=\begin{pmatrix}
%\ones{} & \vec{\sigma} & \zeros{} \\
%\zeros{} & \zeros{} & \vec{\tau}
%\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
%\end{align}

@@

Marginality is an orthogonal concept to uniformity: the distribution of some property is identical \textit{across} all flows. For example, if we determine a distribution of flow sizes across the entire network and assume that this distribution applies \textit{independent of other flow properties}, this is a marginal model for flow size. Marginal behavior is effected by left multiplication by a matrix with uniform columns. This specific example is generated by left-uniformizing the flow size matrix: $B' = (\ones{f \times f}/f)B$.
The traffic models proposed by {\FHC} are all marginal: they propose certain network-wide parametric distributions for start times, flow sizes and the number of flows per node. These models do not account for non-independence between these and other flow properties. We will discuss this limitation and its implications when we analyze our experimental results.

\subsection{Regularized Behavior}

The final form of homogenization we consider, is regularization. This is a restricted form of marginality: we may assume that marginal distributions apply to all the flows in certain groupings. For example, we can provide a distribution of flow sizes on a per-source basis. That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties. This is simply accomplished by left-multiplication by a non-uniform matrix. For example, to make flow behavior uniform on a per source basis, we left-multiply by the ``source-uniformizing'' matrix: $F'=(SS^t/f)F$. Per destination uniformity can be achieved similarly, with left-multiplication by a corresponding destination-uniformizing matrix: $F'=(DD^t/f)F$.
%Although assumptions about uniformity across flows are common, there is little agreement about which notion of uniformity is most appropriate; often it is simply left unspecified. The confusion arises due to the lack of a unique way of associating flows with nodes. Each node can equally viably be associated with its source or its destination. Which association is more valid? Consider a web request: should the behavior of the request be determined by what the client typically does? Or should the requests that the server typically receives determine behavior? There is no armchair answer: only analysis of real data can give us the answer. The matrix model provides an analytical framework for doing so.
The matrix model has the benefit of making the uniformization of behaviors completely explicit. It also allows us to easily define intermediate choices. If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
\begin{align}
F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
\end{align}
The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors. Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.

@@

%\subsection{Uniformity}

%% TODO: differentiate between restrictive versions of models and transformative versions.
%% TODO: distinguish uniform, marginal and regularized models.

%\subsection{Marginality}

%Most approaches to traffic modeling that go beyond uniformity assumptions, instead describe marginal properties of various aspects of traffic behavior. For example, {\FHC}~\cite{Hernandez06:wlan-traffic} provide highly convincing parametric models for six marginal properties of wireless network traffic: session (node) arrivals follow a time-varying Poisson process; per-session flow inter-arrival times are Lognormally distributed; flows per session (node) follow a BiPareto distribution; flow sizes follow a BiPareto distribution. This is certainly much greater detail than simply assuming that all nodes and flows are identical.

@@

\section{Experimental Methodology}
\label{sec:methodology}

@@

\begin{table*}
\begin{center}
\small
\begin{tabular}{|c|c|l|}
\multicolumn{1}{c}{\textbf{model}} &
\multicolumn{1}{c}{\textbf{transformation}} &
\multicolumn{1}{c}{\textbf{description}} \\
\hline
{\footnotesize{GMM}} & --- & the general matrix model \\\hline\hline
time uniform & $A'=A(\ones{d_t \times d_t}/d_t)$ & all start times are equally likely \\\hline
node uniform & $S'=S(\ones{n \times n}/n)$~~~~$D'=D(\ones{n \times n}/n)$ & nodes behave stochastically identically with respect to flows \\\hline
full uniform & \textit{all uniform transformations} & applies all the uniformity assumptions of the above three models \\\hline\hline
time marginal & $A'=(\ones{f \times f}/f)A$ & start time independently chosen from overall distribution of start times \\\hline
size marginal & $B'=(\ones{f \times f}/f)B$ & flow size (bytes) is chosen independently of other properties \\\hline
node marginal & $S'=(\ones{f \times f}/f)S$~~~~$D'=(\ones{f \times f}/f)D$ & nodes behave distinctly, but behavior is independent of flows \\\hline
packet marginal & $Z'=(\ones{f \times f}/f)Z$~~~~$V'=(\ones{f \times f}/f)V$ & packet behavior is variable bit-rate, aggregated across all flows \\\hline
full marginal & \textit{all marginal transformations} & applies all the marginality assumptions of the above four models \\\hline\hline
source regular & $F'=(SS^t/f)F$ & flow behaviors are aggregated on a per source-node basis \\\hline
dest regular & $F'=(DD^t/f)F$ & flow behaviors are aggregated on a per destination-node basis \\\hline
mixed regular & $F'=\frac{1}{2f}(SS^t+DD^t)F$ & flow behaviors are aggregated on a mixed source/destination basis \\\hline
\end{tabular}
\caption{Matrix-based traffic models evaluated by paired differential simulation.}
\label{tab:traffic-models}
\end{center}
\vspace{-2em}
\end{table*}

@@

\begin{figure}[b]
\vspace{-1em}
\begin{center}
\includegraphics[width=3.5in]{nodes-flows}%
\vspace{-0.75em}%
\caption{The number of active nodes and flows over time.} 
\label{fig:nodes-flows}
\end{center}
\vspace{-0.8em}
\end{figure}
